---
title: "Image Recognition using Keras - Convolutional Neural Networks"
author: "Arpit Gupta (arpit.gupta2008@gmail.com)"
output:
  html_document:
    df_print: paged
#output: rmarkdown::github_document
---

<style>
    body .main-container {
        max-width: 1200px;  ## change width
    }
</style>

#### Loading Required packages
```{r,message=F, warning=F}
#devtools::install_github('aoles/EBImage')
library(keras)
library(EBImage)

```

#### Read Images - Training images
```{r}

pic1 <- c('p1.jpg', 'p2.jpg', 'p3.jpg', 'p4.jpg', 'p5.jpg',
          'c1.jpg', 'c2.jpg', 'c3.jpg', 'c4.jpg', 'c5.jpg',
          'b1.jpg', 'b2.jpg', 'b3.jpg', 'b4.jpg', 'b5.jpg')
pic1<-paste("./Data/",pic1,sep = "")
train <- list()
for (i in 1:15) {train[[i]] <- readImage(pic1[i])}
```

#### Read Images - Testing images
```{r}
pic2 <- c('p6.jpg', 'c6.jpg', 'b6.jpg')
pic2<-paste("./Data/",pic2,sep = "")
test <- list()
for (i in 1:3) {test[[i]] <- readImage(pic2[i])}
```

#### Explore
```{r}
print(train[[12]])
summary(train[[12]])
display(train[[12]])
plot(train[[12]])

```

#### Plot all Training Images
```{r}
par(mfrow = c(3,5))
for (i in 1:15) plot(train[[i]])
par(mfrow = c(1,1))
```


```{r}
str(train)
```
#### Resize & combine
```{r}
for (i in 1:15) {train[[i]] <- resize(train[[i]], 100, 100)}
for (i in 1:3) {test[[i]] <- resize(test[[i]], 100, 100)}

train <- combine(train)
x <- tile(train, 5)
display(x, title='Pictures')

```


```{r}
test <- combine(test)
y <- tile(test, 3)
display(y, title = 'Pics')
```

#### Reorder dimension
```{r}
train <- aperm(train, c(4, 1, 2, 3))
test <- aperm(test, c(4, 1, 2, 3))
str(train)

```

#### Response
```{r}
trainy <- c(0, 0, 0, 0, 0, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2)
testy <- c(0, 1, 2)
```


#### One hot encoding
```{r}
trainLabels <- to_categorical(trainy)
testLabels <- to_categorical(testy)
```

## Keras Model
```{r}
# Building sequential model
model <- keras_model_sequential()

model %>%
         layer_conv_2d(filters = 32, 
                       kernel_size = c(3,3),
                       activation = 'relu',
                       input_shape = c(100,100, 3)) %>%
         layer_conv_2d(filters = 32,
                       kernel_size = c(3,3),
                       activation = 'relu') %>%
         layer_max_pooling_2d(pool_size = c(2,2)) %>%
         layer_dropout(rate = 0.25) %>%
         layer_conv_2d(filters = 64,
                       kernel_size = c(3,3),
                       activation = 'relu') %>%
         layer_conv_2d(filters = 64,
                       kernel_size = c(3,3),
                       activation = 'relu') %>%
         layer_max_pooling_2d(pool_size = c(2,2)) %>%
         layer_dropout(rate = 0.25) %>%
         layer_flatten() %>%
         layer_dense(units = 256, activation = 'relu') %>%
         layer_dropout(rate=0.25) %>%
         layer_dense(units = 3, activation = 'softmax') %>%
         
         compile(loss = 'categorical_crossentropy',
                 optimizer = optimizer_sgd(lr = 0.01,
                                           decay = 1e-6,
                                           momentum = 0.9,
                                           nesterov = T),
                 metrics = c('accuracy'))
summary(model)

```

### Fit model
```{r}

history <- model %>%
         fit(train,
             trainLabels,
             epochs = 60,
             batch_size = 32,
             validation_split = 0.2
             #, validation_data = list(test, testLabels)
             )

```

### Plot the model
```{r}

plot(history)
```

### Evaluation & Prediction - train data
```{r}

model %>% evaluate(train, trainLabels)
pred <- model %>% predict_classes(train)
table(Predicted = pred, Actual = trainy)
```

### Evaluation & Prediction of Probabilities - train data
```{r}

prob <- model %>% predict_proba(train)
cbind(prob, Predicted_class = pred, Actual = trainy)
```

### Evaluation & Prediction - test data
```{r}

model %>% evaluate(test, testLabels)
pred <- model %>% predict_classes(test)
table(Predicted = pred, Actual = testy)
```

### Evaluation & Prediction of Probabilities - test data
```{r}

prob <- model %>% predict_proba(test)
cbind(prob, Predicted_class = pred, Actual = testy)
```

```{r}


```


